### 2008-03-21 trunk/Papers/dl-gre-08/conclusion.tex

Because they compute referring expressions for all individuals in the
domain at once, our algorithms will perform especially strongly in
static settings, such as the generation of descriptions for museum
exhibits, in which the individuals and their properties don't change
much.  However, even in more dynamic settings, our algorithms have a
chance to outperform search algorithms like Dale and Haddock's in the
average case because they can't get stuck in unproductive branches of
the search space.  Nevertheless, one interesting question for future
research is how to adapt bisimulation algorithms that can
incrementally update the bisimulation classes when the model changes.
Similarly, it would be interesting to explore how linguistic
constraints and different attribute orderings can be taken into
account without slowing the algorithms down, and how the algorithms
can be adapted to compute REs for sets.

>>>>

Because they compute referring expressions for all individuals in the
domain at once, our algorithms will perform especially strongly in
static settings, such as the generation of descriptions for museum
exhibits, in which the individuals and their properties don't change
much.  However, even in more dynamic settings, our algorithms have a
chance to outperform search algorithms like Dale and Haddock's in the
average case because they can't get stuck in unproductive branches of
the search space. Nevertheless, one interesting question for future
research is how to incrementally update simulation classes when the
model changes. Similarly, it would be interesting to explore how
linguistic constraints and different attribute orderings can be taken
into account without slowing the algorithms down, and how the
algorithms can be adapted to compute REs for sets. In exploring these
extensions we will be able to draw on a rich body of literature that
considers many variants of simulation algorithms, including for
example incremental simulation algorithms or algorithms for
identifying sets of individuals.

%Nevertheless, one interesting question for future
%research is how to adapt simulation algorithms that can
%incrementally update the simulation classes when the model changes.
%Similarly, it would be interesting to explore how linguistic
%constraints and different attribute orderings can be taken into
%account without slowing the algorithms down, and how the algorithms
%can be adapted to compute REs for sets.

### 2008-03-21 trunk/Papers/dl-gre-08/discussion.tex

Both the \el\ and the \alc\ algorithms took about 15 milliseconds to
compute distinguishing concepts for all 16 individuals in the Viethen
and Dale dataset.\footnote{We measured all runtimes on a MacBook Pro
  (Intel Core 2 Duo, 2.16 GHz) running Java 1.6 beta, and allowed the
  Java VM to warm up, i.e.\ just-in-time compile all bytecode, before
  taking the measurements.}

@@

%\begin{itemize}
%\item there's a risk to generate a concept that can't be realized,
%  especially for the \alc\ algorithm
%\item everybody except perhaps SPUD has this problem
%\item problem is worse in our case because it is harder to control the
%  order in which atoms and relations are explored
%\item would be interesting for future work to think about how
%  search in the bisim algorithms can be controlled by the realizer
%\end{itemize}

Our GRE algorithms do not guarantee that the concepts they compute can
actually be realized in language.  For example, none of the concepts
our algorithms computed in the Viethen and Dale domain contained an
atom that would commonly be realized as a noun; the property
$\mathsf{drawer}$ is never used because it applies to all individuals
in the domain.  This particular problem could easily be worked around
in a post-processing step.  However, another problem arises from the
massive use of negation in the \alc\ algorithm; it will be hard for
any realizer to find a reasonable way of expressing a concept like
$\neg \exists R.(\neg P \sqcap \neg Q)$ as a smooth noun phrase.
Although we agree with \newcite{deemter01:_gener_refer_expres} and
others that the careful use of negation and disjunction can improve
REs, these connectives must not be overused.  Thus we consider the
concepts computed by the \el\ algorithm ``safer'' with respect to
realization.

@@

One problem that arises in our approach specifically is that the
bisimulation classes algorithms derive some measure of efficiency from
their freedom to build concepts without having to respect any
linguistic constraints.  It seems straightforward, for instance, to
extend Krahmer et al.'s \shortcite{Krahmer2003} approach such that it
only considers subgraphs that can actually be realized, because their
algorithm proceeds by a genuine search for uniquely identifying
subgraphs, and will simply take a different branch of the search if
some subgraph is useless.  This would be harder in our case.  Our
algorithms don't search in the same way; if we disallow certain
refinements of a partition, we have to allow the algorithms to
backtrack and thus jeopardize the worst-case polynomial runtime.
Investigating this interplay between efficiency and linguistic
constraints is an interesting avenue for future research.

>>>>

Both the \el\ and the \alc\ algorithms took about 15 milliseconds to
compute distinguishing formulas for all 16 individuals in the Viethen
and Dale dataset.\footnote{Runtimes are meassured on a MacBook Pro
  (Intel Core 2 Duo, 2.16 GHz) running Java 1.6 beta, and we allowed the
  Java VM to warm up, i.e., just-in-time compile all bytecode, before
  taking the measurements.}

@@

%\begin{itemize}
%\item there's a risk to generate a formula that can't be realized,
%  especially for the \alc\ algorithm
%\item everybody except perhaps SPUD has this problem
%\item problem is worse in our case because it is harder to control the
%  order in which atoms and relations are explored
%\item would be interesting for future work to think about how
%  search in the bisim algorithms can be controlled by the realizer
%\end{itemize}

Our GRE algorithms do not guarantee that the formula they compute can
actually be realized in language.  For example, none of the formulas
our algorithms computed in the Viethen and Dale domain contained an
atom that would commonly be realized as a noun; the property
$\mathsf{drawer}$ is never used because it applies to all individuals
in the domain.  This particular problem could easily be worked around
in a post-processing step.  However, another problem arises from the
massive use of negation in the \alc\ algorithm; it will be hard for
any realizer to find a reasonable way of expressing a formula like
$\neg \exists R.(\neg P \sqcap \neg Q)$ as a smooth noun phrase.
Although we agree with \newcite{deemter01:_gener_refer_expres} and
others that the careful use of negation and disjunction can improve
REs, these connectives must not be overused.  Thus we consider the
formulas computed by the \el\ algorithm ``safer'' with respect to
realization.

@@

One problem that arises in our approach specifically is that the
bisimulation classes algorithms derive some measure of efficiency from
their freedom to build formulas without having to respect any
linguistic constraints.  It seems straightforward, for instance, to
extend Krahmer et al.'s \shortcite{Krahmer2003} approach such that it
only considers subgraphs that can actually be realized, because their
algorithm proceeds by a genuine search for uniquely identifying
subgraphs, and will simply take a different branch of the search if
some subgraph is useless.  This would be harder in our case.  Our
algorithms don't search in the same way; if we disallow certain
refinements of a partition, we have to allow the algorithms to
backtrack and thus jeopardize the worst-case polynomial runtime.
Investigating this interplay between efficiency and linguistic
constraints is an interesting avenue for future research.

### 2008-03-21 trunk/Papers/dl-gre-08/gre.tex

For the rest of this paper, we assume that we are generating a
singular RE, i.e.\ the target referent $A$ will generally be a
singleton set.  In this case, we will only be able to generate a
formula that denotes exactly $A = \{a\}$ (i.e., a RE that uniquely
refers to $a$) if there is no other individual $b$ to which $a$ is
similar; otherwise, any formula that is satisfied by $a$ is also
satisfied by $b$.  Conversely, if we know that $a$ is not similar to
any other individual, then there is a concept that is satisfied by $a$
and not by anything else; this concept can serve as a unique singular
RE.  In other words, we can reduce the $\gL$-GRE problem for a given
model to the problem of computing the $\gL$-similarity sets of this
model.

>>>>

For the rest of this paper, we assume that we are generating a
singular RE, i.e.\ the target referent $A$ will generally be a
singleton set.  In this case, we will only be able to generate a
formula that denotes exactly $A = \{a\}$ (i.e., a RE that uniquely
refers to $a$) if there is no other individual $b$ to which $a$ is
similar; otherwise, any formula that is satisfied by $a$ is also
satisfied by $b$.  Conversely, if we know that $a$ is not similar to
any other individual, then there is a formula that is satisfied by $a$
and not by anything else; this formula can serve as a unique singular
RE.  In other words, we can reduce the $\gL$-GRE problem for a given
model to the problem of computing the $\gL$-similarity sets of this
model.

### 2008-03-21 trunk/Papers/dl-gre-08/introduction.tex

Referring expressions involving relations, in particular, have
received increasing attention recently, particularly in the context of
spatial referring expressions in situated generation (e.g.\
\cite{kelleher06:_increm_gener_of_spatial_refer}), where it seems
particularly natural to use expressions such as ``the book on the
table''.  However, the classical algorithm by
\newcite{dale91:_gener_refer_expres_invol_relat} was recently shown to
be unable to generate satisfying REs in practice
\cite{viethen06:_algor_for_gener_refer_expres}.  Furthermore, the Dale
\& Haddock algorithm and most of its successors
\cite{Krahmer2003,kelleher06:_increm_gener_of_spatial_refer} are
vulnerable to the problem of ``infinite regress'', where the algorithm
jumps back and forth between generating descriptions for two related
individuals infinitely, as in ``the book on the table which supports a
book on the table \ldots''.

In this paper, we propose to view the problem of generating referring
expressions (REs) as the problem of computing a formula of description
logic (DL) that denotes exactly the set of individuals that we want to
refer to.  This very natural idea has been mentioned in passing before
\cite{Krahmer2003,gardent07:_gener_bridg_defin_descr}; however, we
take it one step further by proposing DL as an interlingua for
comparing the REs produced by different approaches to GRE.  In this
way, we can organize existing GRE approaches in an expressiveness
hierarchy.  For instance, the classical Dale and Reiter algorithms
compute purely conjunctive formulas;
\newcite{deemter01:_gener_refer_expres} extends this language by
adding the other propositional connectives, whereas
\newcite{dale91:_gener_refer_expres_invol_relat} extends it by
allowing existential quantification.

>>>>

Referring expressions involving relations, in particular, have
received increasing attention recently, particularly in the context of
spatial referring expressions in situated generation (e.g.\
\cite{kelleher06:_increm_gener_of_spatial_refer}), where it seems
particularly natural to use expressions such as ``the book on the
table''.  However, the classical algorithm by
\newcite{dale91:_gener_refer_expres_invol_relat} was recently shown to
be unable to generate satisfying REs in practice
\cite{viethen06:_algor_for_gener_refer_expres}.  Furthermore, the Dale
and Haddock algorithm and most of its successors
\cite{Krahmer2003,kelleher06:_increm_gener_of_spatial_refer} are
vulnerable to the problem of ``infinite regress'', where the algorithm
jumps back and forth between generating descriptions for two related
individuals infinitely, as in ``the book on the table which supports a
book on the table \ldots''.

In this paper, we propose to view GRE as the problem of computing a
formula of description logic (DL) that denotes exactly the set of
individuals that we want to refer to.  This very natural idea has been
mentioned in passing before
\cite{Krahmer2003,gardent07:_gener_bridg_defin_descr}; however, we
take it one step further by proposing DL as an interlingua for
comparing the referring expressions (REs) produced by different
approaches to GRE.  In this way, we can organize existing GRE
approaches in an expressiveness hierarchy.  For instance, the
classical Dale and Reiter algorithms compute purely conjunctive
formulas;
\newcite{deemter01:_gener_refer_expres} extends this language by
adding the other propositional connectives, whereas
\newcite{dale91:_gener_refer_expres_invol_relat} extends it by
allowing existential quantification.
